Did WW2 Produce an Equivalent to a Modern Computer?

### 2017-05-03 Draft 2/Section 4 - Zuse's Z3.tex
Improved introduction wording
Conditional branching in Z3
Not quite finished, description of z=j to do

In this section I shall give a broad functional overview of the Z3. I shall explore primarily the function and usage of the machine, the aspects that would be visible to the programmer. I shall not go into detail about the operation of the machine itself unless it directly influences the machine's usage. All of the description of the machine was derived from \cite{rojas1997z3architecture}.

@@

The arithmetical operations perform as follows:
\begin{gather*}
\begin{aligned}
&Ls_1:\quad	&R_1 &= R_1 \times R_2 \\
&Ls_2:\quad	&R_1 &= R_1 \div R_2 \\
&Lm:\quad	&R_1 &= R_1 + R_2 \\
&Li:\quad	&R_1 &= R_1 - R_2 \\
&Lw:\quad	&R_1 &= \sqrt{R_1} \\
\end{aligned}
\end{gather*}

\subsection{Universality in the Z3}

@@

At first glance, it is unclear that the Z3 can be made into a universal machine due to the fact that it allows for neither conditional branching\footnote{The ability to perform different operations depending on the state of a stored value} nor indirect addressing\footnote{The ability to refer to values stored in memory when the address of that memory is not specifically referred to in the program but is rather obtained from the value of some variable currently in memory}. However,

@@

@@

\begin{equation*}
\mathrm{if\ (}z = i\mathrm{)\ then\ } t = 0 \mathrm{\ else\ } t = 1
\end{equation*}

\begin{gather*}
\begin{aligned}
d &= z - i \\
f &= d \cdot d \\
f &= d - e \\
g &= d\ /\ f \\ \\
t &= (2^{16} + g) - 2^{16} \\
\end{aligned}
\end{gather*}

>>>>

In this section I shall give a broad functional overview of the Z3. I shall explore primarily the function and usage of the machine, the aspects that would be visible to the programmer. I shall not go into detail about the operation of the machine itself unless it directly influences the machine's usage. The description of the machine was derived from \cite{rojas1997z3architecture}.

@@

The arithmetical operations perform as follows:
\begin{gather*}
\begin{aligned}
&Ls_1:\quad	&R_1 &= R_1 \times R_2 \\
&Ls_2:\quad	&R_1 &= R_1 \div R_2 \\
&Lm:\quad	&R_1 &= R_1 + R_2 \\
&Li:\quad	&R_1 &= R_1 - R_2 \\
&Lw:\quad	&R_1 &= \sqrt{R_1} \\
\end{aligned}
\end{gather*}
For each operation, the 
\subsection{Universality in the Z3}

@@

At first glance, it is unclear that the Z3 can be made into a universal machine due to the fact that it allows for neither conditional branching\footnote{The ability to perform different operations depending on the state of a stored value} nor indirect addressing\footnote{The ability to refer to values stored in memory when the address of that memory is not specifically referred to in the program but is rather obtained from the value of some variable currently in memory}. However, through use of the tool available to the programmer these features can be emulated, and a Turing Machine can be simulated.

@@

Firstly and fundamentally we shall construct conditional branches using a technique shown in \cite{ibarra1983intdivision} \textit{TODO: Check this ref} which allows the implementation of a form of case statement as found in modern programming languages such as JavaScript. Such an operation performs one of a set of different operations depending on the value of a given variable.

Given a positive integer variable $z$, the variable on which the outcome of the statement depends and the set of operations to be performed depending on the value of $z$, $P_1, ... , P_n$, we can produce a sequence of operations that will allow only the operation $P_z$ to perform. Due to the nature of the Z3 requiring a full sequence of commands to be executed without exception, every one of these operations will be evaluated; however, only operation $P_z$ will be allowed to actually affect memory.

At the start of the switch statement, a sequence of commands check if $z = j$ - if so the flag variable $t$ is set to $0$ (otherwise it is set to $1$). Next, for each operation $P_j$ (which will be of the form $a = b\ \{op\}\ c$), we will write in the program a sequence of commands that, if $t$ is $0$, write the value of $b\ \{op\}\ c$ to $a$. These statements are shown below:

@@

Or, for clarity, $a = (a \cdot t) + (1 - t)(b\ \{op\}\ c)$. If $t = 1$, $v$ is set to $t$; otherwise it is set to $0$. Furthermore, if $t = 0$, $u$ is set to the outcome of the operation; otherwise it is set to $0$ again. Therefore, if $t = 0$, $a$ is set to the outcome of the operation; otherwise it is set back to its original value. As such, after this transformation has been put in place for every $P$ operation, and a method to set $t$ at the start of the case statement has been implemented, every new command in this program can be executed - most superfluously - in order to only use the $z$-th command. Next we must produce a set of operations that will perform the computation
\begin{equation*}
\mathrm{if\ (}z = i\mathrm{)\ then\ } t = 0 \mathrm{\ else\ } t = 1
\end{equation*}

This is done using the following piece of code, where $e$ has been pre-set to a value much less than 1 (e.g. 1/1000).

\begin{gather*}
\begin{aligned}
d &= z - i \\
d &= d \cdot d \\
f &= d - e \\
g &= d\ /\ f \\ \\
t &= (2^{16} + g) - 2^{16} \\
\end{aligned}
\end{gather*}

The variable $d$ is only zero if $z = i$. If not, $d$ is set to a positive value, and $g$ is set to the ratio of $d$ and a value slightly smaller than $d$ - $1$ plus a small fractional part.
